Theory of Phonon Polaritons in Polar Dielectrics

### 2018-08-06 paper.tex
added some references

Phonon polaritons, hybrid quasiparticles of photons and optical phonons, offer great promise for deeply sub-diffractional control of electromagnetic fields at mid-IR and THz frequencies. These polaritons share many features in common with plasmon-polaritons in conductors, a heavily well-studied topic. In recent years, it has been shown that phonon polaritons offer the ability to confine light to volumes over $10^6$ times smaller than that of a photon in free-space at the diffraction limit \cite{caldwell2013low,xu2014mid,caldwell2014sub,dai2014tunable,tomadin2015accessing,yoxall2015direct,li2015hyperbolic,dai2015subdiffractional,dai2015graphene,caldwell2015low,li2016reversible,Basov:2016,basov2017towards,low2017polaritons,giles2017ultra}. Due to this remarkable confinement and their relatively high lifetimes of around picoseconds, phonon polaritons open new opportunities for vibrational spectroscopy, for radiative heat transfer \cite{hillenbrand2002phonon}, and for control of dynamics in quantum emitters \cite{kumar2015tunable,rivera2017making}. Typically, the core features of phonon polaritons, such as mode shape, confinement, and propagation characteristics, can be understood through the properties of the Lorentz oscillator dielectric function, which has been successfully applied to model experiments involving the launching and propagation of these polaritons.

@@

To address these questions an \emph{ab initio} framework to predict the phonon contribution to the dielectric tensor in arbitrary settings is essential. Methodologies exist to calculate the local permittivity contribution from phonons in a 3D bulk, as in \cite{Gonze1997dynamical}, and are implemented in \emph{ab initio} materials physics software packages\cite{abinit1,abinit2,abinit3}. Here, we develop a unified 
theoretical framework to analyze phonon-polaritons in materials, which can address effects of spatial dispersion as well as effects that emerge from arbitrary dimensionality. Specifically, we develop a means to arrive at the dielectric function associated with optical phonons in polar materials of arbitrary dimensionality. We pay close attention to develop a description of the optical response that shows clearly the parallels to the description of plasmonic response from first principles. Importantly, from this theoretical framework, we not only accurately predict surface phonon polaritons in thick materials where a bulk permittivity is valid but also provide tools to predict phonon polariton dispersion in monolayers and anticipate when spatially nonlocal corrections to the phonon contribution to the dielectric function become relevant. We apply this formalism to calculate the dielectric function of silicon carbide and hexagonal boron nitride from first principles, and show that the results agree well with measurements. Finally, we develop the theory of phonon polaritons in a 2D polar material, specifying the dielectric function in terms of \emph{ab initio} parameters, and then finding a universal form for the dispersion relation of 2D phonon polaritons.

@@

%where $\Gamma$ is a positive infinitesimal, and it is to be understood that rigorously, the limit $\Gamma \rightarrow 0$ is to be taken. In a situation where there is phonon dissipation, typically due to electron-phonon and phonon-phonon scattering, we make take a relaxation-time approximation and interpret $\Gamma$ as the finite relaxation time. 

@@

From this equation, let us consider an extremely simplified limit. Assume that the permittivity is isotropic. Further assume that the site-specific masses collapse to a single effective mass $M^*$, and that the site-specific and tensorial Born effective charges collapse to a single effective charge $Q^*$. Finally, assume that the only phonon with oscillator strength (non-zero $\mathbf{S}$) is the transverse optical phonon at frequency $\omega_{TO}$. In this case, one arises at a dielectric function 

@@

\section{Application to bulk semiconductor systems}
% calculation basics
To demonstrate the practical use of this theoretical framework, we present an example calculation of the dielectric function for both hBN and the 4H polytype of silicon carbide (SiC), as is shown in \ref{fig:phonons}(a). 
Phonon bandstructures for SiC and hBN were calculated within density functional perturbation
theory (DFPT) as implemented in the ABINIT package\cite{abinit1,abinit2,abinit3,Gonze1997dynamical,Hamann2005metric}, with electric field response included to appropriately capture LO-TO splitting. Initial ground state density functional theory (DFT) properties were predicted using norm-conserving Vanderbilt pseudopotentials of the PBEsol parameterization for SiC, and PBE with a D2 (Grimme) van der Waals correction for hBN \cite{PBEsol,PBE,ONCV_PPs,pseudodojo,grimme}. We calculated the phonon modes, their frequencies (shown in Figure \ref{fig:phonons}(b)), and their eigendisplacements at the gamma point to construct the dielectric function from Equation \ref{eq:pitoepsilon}, using experimental values for the dissipation rates $\Gamma$. We note that while here, we have taken both the SiC bandgap and the SiC optical phonon lifetime from experiments, these quantities may also be obtained \emph{ab initio} by calculating lifetimes associated with electron-phonon scattering and three-phonon decay processes.

>>>>

Phonon polaritons, hybrid quasiparticles of photons and optical phonons, offer great promise for deeply sub-diffractional control of electromagnetic fields at mid-IR and THz frequencies. These polaritons share many features in common with plasmon-polaritons in conductors, a heavily well-studied topic. In recent years, it has been shown that phonon polaritons offer the ability to confine light to volumes over $10^6$ times smaller than that of a photon in free-space at the diffraction limit \cite{caldwell2013low,xu2014mid,caldwell2014sub,dai2014tunable,tomadin2015accessing,yoxall2015direct,li2015hyperbolic,dai2015subdiffractional,dai2015graphene,caldwell2015low,li2016reversible,Basov:2016,basov2017towards,low2017polaritons,giles2017ultra}. Due to this remarkable confinement and their relatively high lifetimes of around picoseconds, phonon polaritons open new opportunities for vibrational spectroscopy, for radiative heat transfer \cite{hillenbrand2002phonon}, and for control of dynamics in quantum emitters \cite{kumar2015tunable,rivera2017making,kurman2018controlling}. Typically, the core features of phonon polaritons, such as mode shape, confinement, and propagation characteristics, can be understood through the properties of the Lorentz oscillator dielectric function, which has been successfully applied to model experiments involving the launching and propagation of these polaritons.

@@

To address these questions an \emph{ab initio} framework to predict the phonon contribution to the dielectric tensor in arbitrary settings is essential. Methodologies exist to calculate the local permittivity contribution from phonons in a 3D bulk, as in \cite{Gonze1997dynamical}, and are implemented in \emph{ab initio} materials physics software packages\cite{abinit1,abinit2,abinit3}. Here, we develop a unified 
theoretical framework to analyze phonon-polaritons in materials which can address effects of spatial dispersion as well as effects that emerge from arbitrary dimensionality. Specifically, we develop a means to arrive at the dielectric function associated with optical phonons in polar materials of arbitrary dimensionality. We pay close attention to develop a description of the optical response that shows clearly the parallels to the description of plasmonic response from first principles. Importantly, from this theoretical framework, we not only accurately predict surface phonon polaritons in thick materials where a bulk permittivity is valid but also provide tools to predict phonon polariton dispersion in monolayers and anticipate when spatially nonlocal corrections to the phonon contribution to the dielectric function become relevant. We apply this formalism to calculate the dielectric function of silicon carbide and hexagonal boron nitride from first principles, and show that the results agree well with measurements. Finally, we develop the theory of phonon polaritons in a 2D polar material, specifying the dielectric function in terms of \emph{ab initio} parameters, and then finding a universal form for the dispersion relation of 2D phonon polaritons.

@@

@@

From this equation, let us consider an highly simplified limiting case. Assume that the permittivity is isotropic. Further assume that the site-specific masses collapse to a single effective mass $M^*$, and that the site-specific and tensorial Born effective charges collapse to a single effective charge $Q^*$. Finally, assume that the only phonon with oscillator strength (non-zero $\mathbf{S}$) is the transverse optical phonon at frequency $\omega_{TO}$. In this case, one arises at a dielectric function 

@@

\section{Application to bulk semiconductor systems}
% calculation basics
To demonstrate the practical use of this theoretical framework, we present an example calculation of the dielectric function for both hBN and the 4H polytype of silicon carbide (SiC), as is shown in \ref{fig:phonons}(a). 
Phonon bandstructures for SiC and hBN were calculated within density functional perturbation
theory (DFPT) as implemented in the ABINIT package\cite{abinit1,abinit2,abinit3,Gonze1997dynamical,Hamann2005metric}, with electric field response included to appropriately capture LO-TO splitting. Initial ground state density functional theory (DFT) properties were predicted using norm-conserving Vanderbilt pseudopotentials of the PBEsol parameterization for SiC, and PBE with a D2 (Grimme) van der Waals correction for hBN \cite{PBEsol,PBE,ONCV_PPs,pseudodojo,grimme}. We calculated the phonon modes, their frequencies (shown in Figure \ref{fig:phonons}(b)), and their eigendisplacements at the gamma point to construct the dielectric function from Equation \ref{eq:pitoepsilon}, using experimental values for the dissipation rates $\Gamma$. We note that while here, we have taken both the SiC bandgap and the SiC optical phonon lifetime from experiments, these quantities may also be obtained \emph{ab initio} by calculating lifetimes associated with electron-phonon scattering \cite{sundararaman2014theoretical,brown2015nonradiative,brown2016ab,narang2017effects} and three-phonon decay processes \cite{srivastava1990physics}.
